---
title: "Causal analysis based on the benchmarking of the Tubingen datasets"
author: Oleksandr Zadorozhnyi (author thanks to Daniela Schkoda for providing the code for the analysis of the ground truth model in the tubingen dataset)
date: "16/10/2023"
output:
  html_document

---
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```
***Description of the dataset***

The Tubingen cause-effect dataset contains the 108 datasets, every with 2 variables. The description of the data is the following:

 the data represented here is a growing database with different data for testing causal detection algorithms. The goal here is to distinguish between cause and effect. We searched for data sets with known ground truth. However, we do not guarantee that all provided ground truths are correct. The datafiles are .txt-files and contain two variables, one is the cause and the other the effect. For every example there exists a description file where you can find the ground truth and how the data was derived.

Note that not always the first column is the cause and the second the effect. This is indicated in a meta-data file. There is also a weighting factor suggested for some pairs which are very similar if you want to calculate the overall performance.


***Setup of the problem***
We have 89 datasets with variables $Var1, Var2$. The goal is to establish 
```{r}
### loading packages
library("liver")
library("qgraph")
library("igraph")
library("bnlearn")
library("dplyr")

list_of_datasets = read.csv("whitelist.csv")

```
### Loading the Tubingen datasets
```{r}
whitelist = as.list(list_of_datasets)
head(list_of_datasets)
# randomly pick 5 datasets for study comparison 
selected = list_of_datasets[sample(nrow(list_of_datasets),5),]
print(selected)
```
### Here we taking a glimpse in the tubingen data

```{r}
# here we extract the selected datasets to perform benchmarking

sel_to_str = as.character(selected)
elem_len= nchar(sel_to_str)
```
# Transcribing the lengths of the array to form the file names
```{r}
result = vector(mode="character", length=5)
for (i in 1:5){
  text_int = "causal_tubingen"
  elem_len = nchar(sel_to_str[i])
  result[i] <- switch(elem_len,
       "1" = paste(text_int,"00",sel_to_str[i],sep = ""),
       "2" = paste(text_int,"0",sel_to_str[i],sep = ""),
       "3" = paste(text_int,"",sel_to_str[i],sep = ""),
)
}

```


```{r}
library(ggplot2)

set.seed(1)

# The data used can be obtained from https://webdav.tuebingen.mpg.de/cause-effect/. Here we collect automatically the names and descriptions. 
filenames <- sprintf("https://webdav.tuebingen.mpg.de/cause-effect/pair%04d.txt", seq(1, 108, 1))
descriptions <- sprintf("https://webdav.tuebingen.mpg.de/cause-effect/pair%04d_des.txt", seq(1, 108, 1))
```


We are considering the datasets which have 2 variables: in the following we use the HSIC criterion to test for independence of the residuals and covariates
```{r}
library("utils")

results <- data.frame(matrix(nrow = 0, ncol = 2))
colnames(results) <- c("pair", "HSIC test")

# dropping the datasets due to 
drop_f = c(17,64:68)
p_val = 0.05
library(krr)
library(dHSIC)
n = 100 

for (i in setdiff(whitelist$whitelist,drop_f)) {
  # Ground truth files

  if (i %in% c(86, 88)) {
    expected = c(1, 2)
  } else {
    description_file  = url(sprintf("https://webdav.tuebingen.mpg.de/cause-effect/pair%04d_des.txt", 1))
    # Pair 86: x to y, 88 x to y
    description1 = paste(readLines(description_file), collapse=" ")
    description1 = gsub(" ", "", description1)
    if (grepl(">[Yy]|[Yy]<", description1)) {
      expected = c(1, 2)
    } else if (grepl(">[Xx]|[Xx]<", description1)) {
      expected = c(2, 1)
    } else {
      print("This shouldn't happen")
    }
    # print(expected)
  }
  
  # data_file <- url(sprintf("https://webdav.tuebingen.mpg.de/cause-effect/pair%04d.txt", i))
    cause_effect_pair = read.table(filenames[i], header=F)
    cause_effect_pair = as.data.frame(as.matrix(cause_effect_pair[1:min(nrow(cause_effect_pair),n),]))
    var1  = as.matrix(cause_effect_pair[,1])
    var2  = as.matrix(cause_effect_pair[,2])
    # LiNGAM
    # Residual analysis based on HSIC test
    print(i)
    # Fit the ridge regression model
    fit_1 = krr(var1, var2, sigma = 1, lambda = 1)
    #predict_1 =  predict(fit_1, xnew =x1_seq )

    fit_2 = krr(var2, var1, sigma = 1, lambda = 1)
    #predict_2 = predict(fit_2,xnew  = x2_seq)
    
    # Calculate the residuals
    residuals_1 = residuals(fit_1, type = "deviance")
    residuals_2 = residuals(fit_2, type = "deviance")
    
    
    hsic_12 = dhsic.test(residuals_1,var1)
    hsic_21 = dhsic.test(residuals_2,var2)
    
   #  print(sprintf("Pair%04d",i))
   #  print(hsic_12$p.value)
   #  print(hsic_21$p.value)
    
    if (hsic_12$p.value<hsic_21$p.value){
      result = c(1,2)
    }
    else if (hsic_21$p.value<hsic_12$p.value){
      result = c(2,1)
    }
    else {
      result = c(1,2)
    }
    # Results
    results[nrow(results) + 1,] <- c(i, all(expected == result))
}

```


***Refererences***

[1] Reference on HSIC

[2] Reference of Pearson independence test (PC algorithm)

[3] Reference of conditional independence tests


